Secondary structure of the SARS-CoV-2 genome is predictive of nucleotide substitution frequency

Version 1

Abstract

Abstract

Accurate estimation of the effects of mutations on SARS-CoV-2 viral fitness can inform public-health responses such as vaccine development and predicting the impact of a new variant; it can also illuminate biological mechanisms including those underlying the emergence of variants of concern1. Recently, Lan et al reported a high-quality model of SARS-CoV-2 secondary structure and its underlying dimethyl sulfate (DMS) reactivity data2. I investigated whether secondary structure can explain some variability in the frequency of observing different nucleotide substitutions across millions of patient sequences in the SARS-CoV-2 phylogenetic tree3. Nucleotide basepairing was compared to the estimated “mutational fitness” of substitutions, a measurement of the difference between a substitution’s observed and expected frequency that is correlated with other estimates of viral fitness4. This comparison revealed that secondary structure is often predictive of substitution frequency, with significant decreases in substitution frequencies at basepaired positions. Focusing on the mutational fitness of C→T, the most common type of substitution, I describe C→T substitutions at basepaired positions that characterize major SARS-CoV-2 variants; such mutations may have a greater impact on fitness than appreciated when considering substitution frequency alone.

Introduction

While investigating the significance of the substitution C29095T, detected in a familial cluster of SARS-CoV-2 infections5, I hypothesized that this synonymous substitution reflected the high frequency of C→T substitution during the pandemic6. Specifically, frequent C29095T substitution had previously complicated attempts to infer recombinant genomes7. Preliminary investigation of C29095T revealed that it was the fourth most frequent C→T substitution; C29095T occurrs almost seven times as often as a typical C→T substitution4. While there was no clear reason for the selection of this synonymous substitution, C29095 was found to be unpaired in a secondary structure model8. I hypothesized that deamination may be more frequent for unpaired cytosine residues. This was supported by previous analysis with a resolution of ∼300 nucleotides9. To determine whether secondary structure was in fact correlated with mutation frequency at single-nucleotide resolution, the data set reported by Lan et al2 was compared to the mutational fitness estimates reported by Bloom and Neher4. Note that “mutational fitness” is not a measurement of viral fitness per se; rather, it is an estimate made assuming that the expected frequencies of neutral mutations are determined only by the type of substitution (with C→T being much more frequent than all other types of substitutions).

The data compared in this study consisted of estimated mutational fitness for the SARS-CoV-2 genome as reported by Bloom and Neher4 (Supplementary Data ntmut_fitness_all.csv and nt_fitness.csv) as well as population-averaged dimethyl sulfate (DMS) reactivities for SARS-CoV-2-infected Huh7 cells and the corresponding secondary structure model reported by Lan et al2 (Supplementary Data 7 and 8). Note that the estimated mutational fitness is logarithmically related to the ratio of the observed and expected number of occurrences of a nucleotide substitution, with large and asymmetric differences in the frequencies of different types of synonymous substitutions6. Additionally, note that DMS data was obtained in experiments using the WA1 strain in Lineage A, which differs from the more common Lineage B at 3 positions and could have different secondary structure. I focused on the most common types of nucleotide substitutions: those comprising approximately 5% or more of total substitutions.

Results

There was a significant increase in synonymous substitution frequencies at unpaired positions for C→T, G→T, C→A, and T→C, but not for A→G or G→A (p < 0.05; Tukey’s range test with Bonferroni correction; A→T, G→C, and C→G were also significant in an unplanned analysis of all substitution types). For all substitution types with significant differences, unpaired substitution frequencies were higher than basepaired substitution frequencies. The largest effects were observed for C→T and G→T (Figure 1). In this secondary structure model, there is basepairing for 60% and 73% of C and G positions, respectively (limited to those covered in the mutational fitness analysis). Median estimated mutational fitness for synonymous C→T and G→T at unpaired positions are higher than at basepaired positions by 1.46 and 1.36, respectively. Expressed in terms of substitution frequency rather than mutational fitness, the frequency of synonymous C→T and G→T substitution is about four times higher at unpaired positions than at basepaired positions. Together, this demonstrates a meaningful impact of secondary structure on substitution frequencies.

Basepairing is predictive of synonymous substitution frequency.

Distribution of frequencies of synonymous substitutions for the most common substitutions (each approximately corresponding to 5% or more of observed substitutions), expressed as the estimated mutational fitness, which is a logarithmic comparison of the observed versus the expected number of occurrences of each type of substitution in the SARS-CoV-2 phylogenetic tree4. Distributions are grouped by substitution type and whether or not positions are basepaired in a full-genome secondary structure of SARS-CoV-2 in Huh7 cells2.

Basepairing in the secondary structure model appears to be more predictive of estimated mutational fitness than average DMS reactivity, with correlation coefficients of 0.59 and 0.45, respectively, for C→T substitutions (point biserial and Spearman correlation coefficients). Correlation coefficients remain significant, but are reduced (0.18 and 0.13) when considering nonsynonymous mutations (Figure 2, left), consistent with larger and often negative effects of nonsynonymous mutations on viral fitness4. However, DMS reactivity is more correlated with estimated mutational fitness than basepairing when analysis is limited to positions with detectable DMS reactivity (excluding the sites plotted at the minimum measured value of 0.00012). No major difference in this trend was observed across the SARS-CoV-2 genome. As a first-order approximation, two constants were calculated to equalize median mutational fitness for synonymous substitutions at basepaired, unpaired, and all positions. An “adjusted mutational fitness” can then be calculated for C→T substitutions by incrementing mutational fitness by +0.32 at basepaired position and by –1.14 at unpaired positions (results were similar when considering fourfold degenerate positions rather than all synonymous substitutions). Scatter plots compare DMS reactivity to estimated mutational fitness at positions with nonsynonymous C→T substitutions before and after applying this coarse adjustment (Figure 2, left and right).

Estimated mutational fitness correlates with secondary structure for nonsynonymous C→T substitutions.

Scatter plots compare mutational fitness to average DMS reactivity for positions with potential nonsynonymous C→T substitutions. The minimum observed DMS reactivity value is assigned to positions lacking data. Points are colored by basepairing in the full genome secondary structure model. Nonsynonymous C→T substitutions at basepaired positions are highlighted which rank highly for mutational fitness and characterize major SARS-CoV-2 lineages. Synonymous C29095T at an unpaired position is also highlighted. Left: Estimated mutational fitness based only on observed versus expected occurrences of C→T at each position. Right: Mutational fitness adjusted by constants derived from the medians of mutational fitness for synonymous substitutions at basepaired, unpaired, and all potential C→T positions.

For a preliminary estimate of whether nonsynonymous C→T substitutions at basepaired positions are prone to underestimation of mutational fitness, I tested the hypothesis that C→T having highly ranked fitness at basepaired positions would be mutations that characterize significant SARS-CoV-2 lineages (arbitrarily defined as having 5% prevalence in the one-week average of global sequences on the CoV-Spectrum website10 at any time during the pandemic). This was the case for 6 of the top 15 C→T substitutions at basepaired positions; their encoded mutations are shown in Figure 2. Top-ranked mutations characterize Omicron BA.1, one of the first recognized recombinant lineages XB, Gamma P.1, and the current fast-growing lineage JN.1.7. Half of these mutations have relatively high DMS reactivity for basepaired positions and half have very low DMS reactivity. By comparison, the synonymous substitution C29095T at an unpaired position has very high estimated mutational fitness and DMS reactivity. Despite having a higher median estimated mutational fitness (1.41 vs. 1.10), only 3 of the top 15 nonsynonymous C→T at unpaired positions define major lineages (BQ.1.1, JN.1.8.1, and BA.2.86.1).

Of particular note is C22227T at a basepaired position encoding the spike A222V mutation. This was one mutation that characterized the B.1.177 lineage, and it was unclear whether it conferred any fitness advantage11. Further investigation as well as its recurrence in the major Delta sublineage AY.4.2 provided additional evidence of an increase in viral fitness and suggested molecular mechanisms12. Here, I focus on C→T substitutions for comparison to DMS reactivity data, but I also note that top-ranked G→T substitutions at basepaired positions are rich in mutations to ORF3a and also include mutations that characterize variants of concern, such as nucleocapsid D377Y in Delta. Lastly, note that, following the coarse adjustment for basepairing inferred from synonymous substitutions, nonsynonymous C→T substitutions characterizing major variants now have some of the highest estimated mutational fitnesses for C→T substitutions (Figure 2, right).

Discussion

This analysis shows that it is informative to combine apparent viral fitness, inferred from massive sequencing of SARS-CoV-2 during the pandemic, with accurate secondary structure measurements. It is important to remember that apparent “mutational fitness” results from a combination of the rates at which diversity is generated and the subsequent selection processes. Importantly, genome secondary structure can impact both. However, even the unprecedented density of sampling SARS-CoV-2 genomes has been insufficient to reliably infer fitness impacts of single mutations more directly from dynamics subsequent to occurrences in the SARS-CoV-2 phylogenetic tree4,13. Further investigation into phenomena reported here, such as the lack of apparent secondary structure dependence for A→G and G→A substitutions, could inform investigation of underlying mutation mechanisms. I suggest that secondary structure, along with other data correlating with substitution frequency, can be used to refine estimates of mutational fitness.

More sophisticated analysis can incorporate structural heterogeneity2 as well as local sequence context14. Furthermore, additional measurements of secondary structure for genomes of new variants or modeling may reveal significant changes to secondary structure since 2020. For the spike protein, the correlation between estimated mutational fitness and pseudovirus entry quantified by deep mutational scanning serves as one metric that can be used to optimize models15. However, it is critical to evaluate uncertainty in any model estimating fitness of a new variant. To this end, initial estimates can be refined by rapid in vitro characterization and continued genomic surveillance.

eLife Assessment

This short manuscript uses mutation counts in phylogenies of millions of SARS-CoV-2 genomes to show that mutation rates systematically differ between regions that are paired or unpaired in the predicted RNA secondary structure of the viral genome. Such an effect of pairing state is not unexpected, but its systematic demonstration using millions of viral genomes is valuable and convincing.

Reviewer #1 (Public Review):

Summary:

This very short paper shows a greater likelihood of C->U substitutions at sites predicted to be unpaired in the SARS-CoV-2 RNA genome, using previously published observational data on mutation frequencies in SARS-CoV-2 (Bloom and Neher, 2023).

General comments:

A preference for unpaired bases as a target for APOBEC-induced mutations has been demonstrated previously in functional studies so the finding is not entirely surprising. This of course assumes that A3A or other APOBEC is actually the cause of the majority of C->U changes observed in SARS-CoV-2 sequences.

I'm not sure why the authors did not use the published mutation frequency data to investigate other potential influences on editing frequencies, such as 5' and 3' base contexts. The analysis did not contribute any insights into the potential mechanisms underlying the greater frequency of C->U (or G->U) substitutions in the SARS-CoV-2 genome.

Reviewer #2 (Public Review):

Hensel investigated the implications of SARS-CoV-2 RNA secondary structure in synonymous and nonsynonymous mutation frequency. The analysis integrated estimates of mutational fitness generated by Bloom and Neher (from publicly available patient sequences) and a population-averaged model of RNA basepairing from Lan et al (from DMS mutational profiling with sequencing, DMS-MaPseq).

The results show that base-pairing limits the frequency of some synonymous substitutions (including the most common CT), but not all: GA and AG substitutions seem unaffected by base-pairing.

The author then addressed nonsynonymous CT substitutions at base-paired positions. While there is still a generally higher estimated mutational fitness at unpaired positions, they propose a coarse adjustment to disentangle base-pairing from inherent mutational fitness at a given position. This adjustment reveals that nonsynonymous substitutions at base-paired positions, which define major variants, have higher mutational fitness.

Overall, this manuscript highlights the importance of considering RNA secondary structure in viral evolution studies.

The conclusions of this work are generally well supported by the data presented. Particularly, the author acknowledges most limitations of the analyses, and addresses them. Even though no new sequencing results were generated, the author used available data generated from the analysis of roughly seven million sequenced patient samples. Finally, the author discusses ways to improve the current available models.

There are a number of limitations of this work that should be highlighted, specifically in regard to the secondary structure data used in this paper. The Lan et al. dataset was generated using a multiplicity of infection (MOI) of 0.05, 24 hours post-infection (h.p.i.). At such a low MOI and late timepoint, viral replication is not synchronous and sequencing artifacts might be generated by cell debris and viral RNA degradation, therefore impacting the population-averaged results. In addition, the nonsynonymous base-paired positions in Figure 2 have relatively high population-averaged DMS reactivity, which suggests those positions are dynamic. Therefore, the proposed adjustment could result in an incorrect estimation of their inherent mutational fitness.

Additionally, like all such RNA probing experiments within cells, it remains difficult to deconvolve DMS/SHAPE low reactivity with RNA accessibility (e.g. from protein binding).

This work presents clear methods and an easy-to-access bioinformatic pipeline, which can be applied to other RNA viruses. Of note, it can be readily implemented in existing datasets. Finally, this study raises novel mechanistic questions on how mutational fitness is not correlated to secondary structure in the same way for every substitution.

Overall, this work highlights the importance of studying mutational fitness beyond an immune evasion perspective. On the other hand, it also adds to the viral intrinsic constraints to immune evasion.

Author response:

The following is the authors’ response to the original reviews.

Reviewer 1 Public Review:

Summary

This very short paper shows a greater likelihood of C->U substitutions at sites predicted to be unpaired in the SARS-CoV-2 RNA genome, using previously published observational data on mutation frequencies in SARS-CoV-2 (Bloom and Neher, 2023).

General comments

A preference for unpaired bases as a target for APOBEC-induced mutations has been demonstrated previously in functional studies so the finding is not entirely surprising. This of course assumes that A3A or other APOBEC is actually the cause of the majority of C->U changes observed in SARS-CoV-2 sequences.

I'm not sure why the authors did not use the published mutation frequency data to investigate other potential influences on editing frequencies, such as 5' and 3' base contexts. The analysis did not contribute any insights into the potential mechanisms underlying the greater frequency of C->U (or G->U) substitutions in the SARS-CoV-2 genome.

I have added additional discussions of mechanisms focusing on the question of whether basepairing bias is primarily driven by secondary structure dependence of underlying mutation rates or by conservation of secondary structure (Discussion lines 178–192) and I added a brief analysis of the 5′ and 3′ contexts of the relationship between being basepaired in a secondary structure model and apparent mutational fitness (Figures S1 and S2, Results lines 85–97). I found that the 5′ context of unpaired, but not paired basepairs influences apparent mutational fitness (preference for 5′ U), and that the is also . Additionally, there is a 3′ preference for G, indicating some CpG suppression. This contrasts to some degree with another analysis based on counting lineage frequencies that may have lacked power to detect relatively small effects (Simmonds mBio 2024).

Reviewer 1 Author recommendations:

There are at least 5 publications describing the mapping/prediction of SARS-CoV-2 RNA secondary structure from 2022-2023 and their predictions are not entirely consistent. Why did the authors only refer to the Lan et al. paper?

I have added comparisons when the Lan et al secondary structure model is replaced by one of two others derived from SHAPE data (Results lines 110–122). Unsurprisingly, similar secondary structure models give similar results and performance is modestly higher for the models from Lan et al. This is consistent with their observations that DMS reactivities performed better as classifiers of SL5 and ORF1 secondary structure (the reason I compared to this secondary structure model and reactivity data set rather than others), but I did not go into detail on this in the revision since there are many differences in methods beyond class of reactivity probe. For example, somewhat stronger correlation for the Vero than the Huh7 dataset in Lan et al could arise from combining data from two replicates, from cell type, or from differences in data analysis methods. It’s also a small difference and cannot be confidently distinguished from noise.

I conducted a preliminary comparison of the performance of DMS and SHAPE data for predicting mutations where DMS data is available, but I opted against including this analysis in the manuscript for the same reasons. Instead, I included in results and discussion comments on how, in general, reactivity data contains information that is predictive of substitution rates that is not captured by binary secondary structure models. I also discuss how multiple data sources can potentially be integrated to more accurately predict the impact of a substitution on fitness (Discussion lines 195–201).

Specific substitutions are referred to as C->T and C29085T for example, but as the genome of SARS-CoV-2 is RNA, and T should be a U.

I agree and I have changed all “T” to “U” in the paper and analysis scripts. The choice of “T” was motivated by what seemed to appear most frequently in papers on SARS-CoV-2 mutational spectra, but “U” is nearly universal in papers on secondary structure and mutation mechanisms, so I agree it makes more sense in this paper.

The C29085T substitution is somewhat non-canonical as it is a single base bulge in a longer duplex section of dsRNA, very unlike the favoured sites for mutation in the Nakata et al paper.

I have added a discussion of Nakata et al ( NAR 2023) ( Introduction lines 29–32). I did not go into this depth in the revision, but the analysis of ~2M patient sequences in Nakata et al also noted a high rate of UUC→UUU substitution, so the UUUC context of C29095 (shared by 3 of the 10 positions highlighted in Nakata et al that had high mutation frequencies with exogenous APOBEC3A expression) could be interesting to investigate further.

High C29095U substitution frequency is indeed somewhat at odds with the results in that work, which found that UC→UU substitutions to be elevated in longer single-stranded regions than the context of C29095U in SARS-CoV-2 secondary structure models (a single unpaired base opposing three unpaired bases in an asymmetric internal loop).

I'm not sure why DMS reactivity is considered a separate variable from pairing likelihood as one informs the other.

The intent here, which was not clear, was to show that a binary basepairing model that uses DMS reactivities as constraints does not capture all of the information available. I have clarified this in as described above discussing information in different reactivy datasets.

The C29095U substitution is also relavent to the consideration of DMS reactivity in addition to the resulting secondary structure model. These are not considered as separate predictors and the reason for showing both is mentioned in the paper: “DMS reactivity was more strongly correlated with estimated mutational fitness than basepairing when analysis was limited to positions with detectable DMS reactivity.” I have clarified this in the revised manuscript and also it is relevant to the discussion of a potential model integrating all available datasets.

Reviewer 2 Public Review:

Hensel investigated the implications of SARS-CoV-2 RNA secondary structure in synonymous and nonsynonymous mutation frequency. The analysis integrated estimates of mutational fitness generated by Bloom and Neher (from publicly available patient sequences) and a population-averaged model of RNA basepairing from Lan et al (from DMS mutational profiling with sequencing, DMS-MaPseq).

The results show that base-pairing limits the frequency of some synonymous substitutions (including the most common CT), but not all: GA and AG substitutions seem unaffected by base-pairing.

The author then addressed nonsynonymous CT substitutions at base-paired positions. While there is still a generally higher estimated mutational fitness at unpaired positions, they propose a coarse adjustment to disentangle base-pairing from inherent mutational fitness at a given position. This adjustment reveals that nonsynonymous substitutions at base-paired positions, which define major variants, have higher mutational fitness.

Overall, this manuscript highlights the importance of considering RNA secondary structure in viral evolution studies.

The conclusions of this work are generally well supported by the data presented. Particularly, the author acknowledges most limitations of the analyses, and addresses them. Even though no new sequencing results were generated, the author used available data generated from the analysis of roughly seven million sequenced patient samples. Finally, the author discusses ways to improve the current available models.

There are a number of limitations of this work that should be highlighted, specifically in regard to the secondary structure data used in this paper. The Lan et al. dataset was generated using a multiplicity of infection (MOI) of 0.05, 24 hours post-infection (h.p.i.). At such a low MOI and late timepoint, viral replication is not synchronous and sequencing artifacts might be generated by cell debris and viral RNA degradation, therefore impacting the population-averaged results. In addition, the nonsynonymous base-paired positions in Figure 2 have relatively high population-averaged DMS reactivity, which suggests those positions are dynamic. Therefore, the proposed adjustment could result in an incorrect estimation of their inherent mutational fitness.

I would go further than this to say that the proposed adjustmentment will usually result in an incorrect estimate. My intent is to propose an improved, but still likely incorrect, estimate by utilizing in vitro data to refine baseline mutation rates in order to obtain improved, but only coarsely adjusted, estimates of mutational fitness. I added a note in the discussion that in vitro reactivities (and, consequently, secondary structure models) may not reflect secondary structures in vivo ( Discussion lines 204–205). I did not go into detail regarding the specific technical considerations mentioned here because they are outside the scope of my expertise.

I am not sure that top-ranked non-synonymous C→U positions have particularly high DMS values after coarse adjustment for basepairing (labeled amino acid mutations in Figure 2). Of the six common mutations used as examples, three have minimum values in the dataset considered (which is processed normalized/filtered data rather than raw data) and three do not have very high DMS reactivity.

However, there is clearly information in base reactivity that is not captured by a binary basepairing model, which is indicated by residual positive correlation between DMS reactivity and mutational fitness after adjustment. I now include a figure demonstrating this for synonymous C→U substitutions as Figure S3, and I have tried to clarify the language throughout the manuscript to make it clear that a more accurate adjustment is possible.

Additionally, like all such RNA probing experiments within cells, it remains difficult to deconvolve DMS/SHAPE low reactivity with RNA accessibility (e.g. from protein binding).

I agree, and in revising this manuscript it was interesting to see that Nakata et al ( discussed above) identified relatively large single-stranded regions with enhanced UC→UU substitution frequencies with exogenous APOBEC3A expression, while C29095U, for example, is a single unpaired base with high DMS reactivity and high empirical C→U substitution frequency (discussed briefly in the introduction of the revised manuscript). Future analyses could consider heterogeneity in secondary structure as well as secondary structures with low heterogeneity where strained conformations could have higher reactivity.

This work presents clear methods and an easy-to-access bioinformatic pipeline, which can be applied to other RNA viruses. Of note, it can be readily implemented in existing datasets. Finally, this study raises novel mechanistic questions on how mutational fitness is not correlated to secondary structure in the same way for every substitution.

Overall, this work highlights the importance of studying mutational fitness beyond an immune evasion perspective. On the other hand, it also adds to the viral intrinsic constraints to immune evasion.

Reviewer 2 Author recommendations:

Even though the experiment was not performed in this manuscript, it would be helpful for the readers if it was briefly explained how secondary structure is inferred from DMS reactivity, as this technique is not broadly used.

It is not objective to refer to the Lan et al. model of RNA structure as "high quality" given the limitations of their experimental approach (low MOI, asynchronous infection, DMS-only, no long-range interactions) and the lack of external validation of the structure of the genome they propose.

I removed “high-quality” from the abstract. Since a result of the paper is that secondary structure correlates with synonymous substitution rates, this is an observation that can be used to retrospectively compare the quality of secondary structure models in this respect. I updated the manuscript to include such a comparison, and did not find a large difference between secondary structure models (Results lines 110–122). I added a discussion of how multiple data sources can potentially be integrated to more accurately predict the impact of a substitution of viral fitness.

I have also added a brief discussion of constraints on how much we can confidently infer from these experiments given limitations of the experimental approach. I note that DMS and SHAPE data provide information that can be combined to make a stronger model, and that predictions can be rapidly tested given observations by Gout (Symonds?) et al that in vitro substitution rates correlate with those observed during the pandemic (Discussion lines 195–201).

Mutational fitness from Bloom & Neher was derived throughout the pandemic, much of which came from a period with the most active surveillance (Delta / Omicron waves). Consequently, these viruses differ from the WA1 strain used by Lan et al. far more than the 3 nt differences between lineage A and B that the author refers to. The following sentence should therefore be revised to avoid misleading the reader:

"Additionally, note that DMS data was obtained in experiments using the WA1 strain in Lineage A, which differs from the more common Lineage B at 3 positions and could have different secondary structure."

Revised:

“Additionally, note that DMS data was obtained in experiments using the WA1 strain in Lineage A, which differs from the more common Lineage B at 3 positions and could have different secondary structure. Furthermore, mutational fitness is estimated from the phylogenetic tree of published sequences (the public UShER tree (Turakhia et al., 2021) additionally curated to filter likely artifacts such as branches with numerous reversions) that are typically far more divergent and subsequently will have somewhat different secondary structures. Since the dataset used for mutational fitness aggregates data across viral clades, my analysis will not capture secondary structure variation between clades or indels and masked sites that were not considered in that analysis (Bloom and Neher, 2023).”

To determine the extent to which the results depend on the single RNA structure model, it would be informative "turn the crank again" on the analysis with one of the other RNA structure datasets for SARS-CoV-2 (though most other datasets suffer from similar problems of asynchronicity of infection).

I have added comparisons when the Lan et al secondary structure model is replaced by one of two others derived from SHAPE data as described above. Also, I conducted preliminary comparisons of underlying DMS and SHAPE reactivity data as described above, but I opted not to include these in the revised manuscript given that methods different beyond the chemical probe used. I also discuss how multiple data sources can potentially be integrated to more accurately predict the impact of a substitution of viral fitness.

In Figure 1 it would be helpful to add the values of the unpaired/basepaired ratios in the plot for clarity.

Furthermore, a similar analysis using the substitution frequency, which strengthens the conclusions, is mentioned in the text, however, it is not shown. It could be shown as part of Figure 1, or as a supplementary figure.

This was a good suggestion since numbers around 1 are not perceived as being very significant. I added the ratio of median unpaired:paired rates to Figure 1, updated the corresponding manuscript text and the figure caption, and note that the numbers are somewhat changed from the first version of my manuscript because of updating to use the most up-to-date mutational fitness estimates.

It is not clear how the two constants were calculated to obtain the "adjusted mutational fitness". It could be shown as part of Figure 2, or as a supplementary figure.

I added dashed lines and arrows to Figure 2 showing median paired/unpaired mutational fitnesses and the adjustment made to normalize to the overall median. I also added Figure S3 showing this for synonymous substitutions, where it is more clear given the lower fraction of mutations with substantial fitness impacts.

Minor comments

Statements like "the current fast-growing lineage JN.1.7" never age well... please revise to state the period of time to which this refers.

Revised:

“…lineage JN.1.7, which had over 20% global prevalence in Spring 2024…”

Also, I checked the list of mutations and the examples given remain in the top 15 ranked basepaired, non-synonymous C→U mutations (BA.2-defining C26060U is added to the list, but I did not update to include this). It replaces C9246U, which was not mentioned in the first version of the manuscript.

Similarly, please provide context for the reader in the phrase: "This was one mutation that characterized the B.1.177 lineage" (e.g. add its early reference as "EU1" and that it predominated in Europe in autumn 2020, prior to the emergence of the Alpha variant).

Revised to add detail:

This was one of the mutations that characterized the B.1.177 lineage. This lineage, also known as EU1, characterized a majority of sequences in Spain in summer 2020 and eventually in several other countries in Europe prior to the emergence of the Alpha variant. However, it was unclear whether or this lineage had higher fitness than other lineages or if A222V specifically conferred a fitness advantage.

"massive sequencing of SARS-CoV-2" - the meaning of the word "massive" is unclear. Revise.

Revised “…millions of patient SARS-CoV-2 sequences published during the pandemic…”

Version 2

Abstract

Abstract

Accurate estimation of the effects of mutations on SARS-CoV-2 viral fitness can inform public-health responses such as vaccine development and predicting the impact of a new variant; it can also illuminate biological mechanisms including those underlying the emergence of variants of concern (Carabelli et al., 2023). Recently, Lan et al. reported a model of SARS-CoV-2 secondary structure and its underlying dimethyl sulfate (DMS) reactivity data (Lan et al., 2022). I investigated whether base reactivities and secondary structure models derived from them can explain some variability in the frequency of observing different nucleotide substitutions across millions of patient sequences in the SARS-CoV-2 phylogenetic tree. Nucleotide basepairing was compared to the estimated “mutational fitness” of substitutions, a measurement of the difference between a substitution’s observed and expected frequency that is correlated with other estimates of viral fitness (Bloom and Neher, 2023). This comparison revealed that secondary structure is often predictive of substitution frequency, with significant decreases in substitution frequencies at basepaired positions. Focusing on the mutational fitness of C→U, the most common type of substitution, I describe C→U substitutions at basepaired positions that characterize major SARS-CoV-2 variants; such mutations may have a greater impact on fitness than appreciated when considering substitution frequency alone.

Introduction

While investigating the significance of the substitution C29095U, detected in a familial cluster of SARS-CoV-2 infections (Chan et al., 2020), I hypothesized that this synonymous substitution reflected the high frequency of C→U substitution during the pandemic (De Maio et al., 2021). Specifically, frequent C29095U substitution had previously complicated attempts to infer recombinant genomes (VanInsberghe et al., 2021). Preliminary investigation revealed that C29095U was the fourth most frequent C→U substitution, occurring almost seven times as often as a typical C→U substitution (Bloom and Neher, 2023). While there was no clear reason for selection of this synonymous substitution, C29095 was found to be unpaired in a secondary structure model (Sun et al., 2021). The 5′ context of C29095 is UC, and additional SARS-CoV-2 secondary structure models that I consulted (see below) agree that C29095 occurs as a single unpaired base in an asymmetrical internal loop, opposite three unpaired bases. Although one study found that exogenous expression of APOBEC3A accelerated C→U substitution in SARS-CoV-2-infected cells in the UC context (Nakata et al., 2023), the most frequently mutated positions were in the context of larger unpaired regions.

I hypothesized that non-enzymatic deamination may be more frequent for unpaired cytosine residues, which was supported by a previous analysis with a resolution of approximately 300 nucleotides (Li et al., 2023) and by a comparison of sequence diversity in early SARS-CoV-2 genomes to an RNA folding prediction (Simmonds, 2020). However, these studies differed in their interpretations of relatively high C→U substitution frequencies at unpaired positions, leaving it unresolved whether this was primarily driven by higher mutation rates at unpaired positions or fitness costs of mutations disrupting secondary structure. To determine whether secondary structure was in fact correlated with mutation frequency at single-nucleotide resolution and to extend this analysis beyond C→U substitutions, the dataset reported by Lan et al. was compared to the mutational fitness estimates reported by Bloom and Neher (Bloom and Neher, 2023). Note that “mutational fitness” is not a measurement of viral fitness per se; rather, it is an estimate made assuming that the expected frequencies of neutral mutations are determined only by the type of substitution (with C→U being much more frequent than all other types of substitutions).

Methods

The data compared in this study consisted of estimated mutational fitness for the SARS-CoV-2 genome as reported by Bloom and Neher (Supplementary Data: ntmut_fitness_all.csv; November 2024 dataset) as well as population-averaged dimethyl sulfate (DMS) reactivities for SARS-CoV-2-infected Huh7 cells and the corresponding secondary structure model reported by Lan et al. (Supplementary Data 7 and 8). Two other secondary structure models based upon SHAPE reactivity data were also used (Huston et al., 2021; Manfredonia et al., 2020). Note that the estimated mutational fitness is logarithmically related to the ratio of the observed and expected number of occurrences of a nucleotide substitution, with large and asymmetric differences in the frequencies of different types of synonymous substitutions (De Maio et al., 2021). Additionally, note that DMS data was obtained in experiments using the WA1 strain in Lineage A, which differs from the more common Lineage B at 3 positions and could have different secondary structure. Furthermore, mutational fitness is estimated from the phylogenetic tree of published sequences (the public UShER tree (Turakhia et al., 2021) additionally curated to filter likely artifacts such as branches with numerous reversions) that are typically far more divergent and subsequently will have somewhat different secondary structures. Since the dataset used for mutational fitness aggregates data across viral clades, my analysis will not capture secondary structure variation between clades or indels and masked sites that were not considered in that analysis (Bloom and Neher, 2023). I focused on the most common types of nucleotide substitutions: those comprising approximately 5% or more of total substitutions.

Results

There was a significant increase in synonymous substitution frequencies at unpaired positions for C→U, G→U, C→A, and U→C, but not for A→G or G→A (p < 0.05; Tukey’s range test with Bonferroni correction; A→U, G→C, and C→G were also significant in an unplanned analysis of all substitution types). For all substitution types with significant differences, unpaired substitution frequencies were higher than basepaired substitution frequencies. The largest effects were observed for C→U and G→U (Figure 1). In this secondary structure model, there is basepairing for 60% and 73% of C and G positions, respectively (limited to those covered in the mutational fitness analysis). The largest impacts on median estimated mutational fitness were observed for synonymous C→U and G→U mutations, with apparent fitness higher at unpaired positions than at paired positions by 1.44 and 1.25, respectively. Expressed in terms of substitution frequency rather than mutational fitness (Figure 1), the frequency of synonymous C→U and G→U substitutions were about four times higher at unpaired positions than at base-paired positions. Together, this demonstrates a meaningful impact of secondary structure on substitution frequencies.

Basepairing is predictive of synonymous substitution frequency.

Distribution of frequencies of synonymous substitutions for the most common substitutions (each approximately corresponding to 5% or more of observed substitutions), expressed as the estimated mutational fitness, which is a logarithmic comparison of the observed versus the expected number of occurrences of each type of substitution in the SARS-CoV-2 phylogenetic tree (Bloom and Neher, 2023). Distributions are grouped by substitution type and whether positions are basepaired in a full-genome secondary structure model of SARS-CoV-2 in Huh7 cells (Lan et al., 2022). The median relative substitution rate (ratio of median rates of substitutions at unpaired and paired positions) is shown above each type of substitution.

Similar results were reported by two studies after this study was first preprinted. One study identified a basepairing dependence for C→U substitutions during serial passage of different SARS-CoV-2 variants and considered mutational fitness estimates from Bloom and Neher for more rare substitution types (Gout et al., 2024). In another analysis that analyzed recurrent substitutions in multiple lineages, little difference was observed overall for C→U substitutions in different 5′ and 3′ contexts (Simmonds, 2024). It was observed that only the few C→U substitutions that appeared with the highest frequencies had a strong preference for 5′ U context.

Figure S1 shows that the degree of this 5′ U preference becomes more clear when limiting analysis to synonymous substitutions and separately analyzing basepaired positions (which show no 5′ preference) and unpaired positions. At unpaired positions, synonymous C→U substitutions with 5′ U context have a higher apparent fitness (0.83 [0.79, 0.89], 95% confidence interval from 1000 bootstrapped samples) than with 5′ G (0.15 [0.01, 0.50]), 5′ A (0.55 [0.45, 0.68]), or 5′ C (0.57 [0.33, 0.69]). Conversely, synonymous C→U substitution rates with 3′ G context were elevated at both unpaired and paired positions (Figure S2), suggesting some CpG suppression with little dependence on basepairing. Notably, the previously mentioned C29095 is both unpaired and appears in the context UCG.

Basepairing in the secondary structure model appeared to be more predictive of estimated mutational fitness than average DMS reactivity, with correlation coefficients of 0.59 [0.55, 0.62] and 0.45 [0.40, 0.49], respectively, for synonymous C→U substitutions (point biserial and Spearman correlation coefficients; 95% confidence intervals from 1000 bootstrapped samples). However, DMS reactivity was more strongly correlated with estimated mutational fitness than base-pairing when analysis was limited to positions with detectable DMS reactivity (0.77 [0.74, 0.79], excluding the sites plotted at the minimum measured value of 0.00012). This suggests that considering structural heterogeneity captured in base reactivity datasets could improve predictions of substitution rates. Correlation coefficients remained significant, but were reduced (0.17 [0.14, 0.20] and 0.13 [0.09, 0.16]) when considering nonsynonymous C→U substitutions (Figure 2, left), consistent with larger and often negative effects of nonsynonymous mutations on viral fitness (Bloom and Neher, 2023).

Estimated mutational fitness correlates with secondary structure for nonsynonymous C→U substitutions.

Scatter plots compare mutational fitness to average DMS reactivity for positions with potential nonsynonymous C→U substitutions. The minimum observed DMS reactivity value is assigned to positions lacking data. Points are colored by basepairing in the full genome secondary structure model. Nonsynonymous C→U substitutions at basepaired positions are highlighted which rank highly for mutational fitness and characterize major SARS-CoV-2 lineages. Synonymous C29095U at an unpaired position is also highlighted. Left: Estimated mutational fitness based only on observed versus expected occurrences of C→U at each position. Dashed lines indicate the median estimated mutational fitness for synonymous substitutions at paired and unpaired positions. Arrows indicate the magnitudes of adjustments made to mutational fitness that result in median fitness of synonymous substitutions at paired (+0.32) and unpaired (−1.13) positions identical to the unadjusted median for all synonymous substitutions (−0.46). Right: Mutational fitness adjusted by constants derived from the medians of mutational fitness for synonymous substitutions at basepaired, unpaired, and all potential C→U positions.

Since basepairing is correlated with substitution rates, the degree of correlation between secondary structure models and substitution rates can be used to compare models that differ in underlying experimental methods and analysis. In light of this, secondary structure models derived from two DMS reactivity datasets (from SARS-CoV-2 isolate WA1 infection of Huh7 and Vero cells) (Lan et al., 2022) and two SHAPE reactivity datasets (SARS-CoV-2 isolates WA1 and Leiden0002 infection of Huh7.5 and Vero cells, respectively) (Huston et al., 2021; Manfredonia et al., 2020) were compared by degree of correlation with estimated mutational fitness. Correlations between basepairing and synonymous C→U substitution rates were highest for models derived from DMS data (0.59 [0.55, 0.62] and 0.63 [0.60, 0.67] for Huh7 and Vero datasets, respectively). Models derived from SHAPE data had modestly lower correlations (0.55 [0.50, 0.59] and 0.50 [0.46, 0.54] for Huh7 and Vero datasets, respectively). However, the methods used to construct each of these four models have difference beyond base reactivity probes, cell type, and viral isolate, so small differences in correlation cannot be attributed to a single factor.

Thus, the high mutation rate of C29095U could be more accurately estimated by considering DMS reactivity than basepairing alone (Figure 2) since it is in the 99th percentile of DMS activity for positions with synonymous C→U substitutions. No major difference in this trend was observed across the SARS-CoV-2 genome. As a first-order approximation, two constants were calculated to equalize median mutational fitness for synonymous substitutions at base-paired, unpaired, and all positions. An “adjusted mutational fitness” can then be calculated for C→U substitutions by incrementing mutational fitness by +0.32 at basepaired position and by −1.13 at unpaired positions (results were similar when considering fourfold degenerate positions rather than all synonymous substitutions). Scatter plots compare DMS reactivity to estimated mutational fitness at positions with nonsynonymous C→U substitutions before and after applying this coarse adjustment (Figure 2, left and right).

This adjustment equalizes median mutational fitness for synonymous C→U substitutions (Figure S3), and remaining correlation between “adjusted mutational fitness” and DMS reactivity (0.18 [0.13, 0.22]) shows that this coarse adjustment could be improved with more complex analysis. However, a binary model of basepairing alone is sufficient to capture much of the variation in synonymous substitution rates and could be extended to other substitution types with rates depending on substitution types that lack reactivity data.

For a preliminary estimate of whether nonsynonymous C→U substitutions at basepaired positions are prone to underestimation of mutational fitness, I tested the hypothesis that C→U having highly ranked fitness at basepaired positions would be mutations that characterize significant SARS-CoV-2 lineages (arbitrarily defined as having 5% prevalence in the one-week average of global sequences on the CoV-Spectrum website (Chen et al., 2022) at any time during the pandemic). This was the case for 6 of the top 15 C→U substitutions at basepaired positions; their encoded mutations are shown in Figure 2. Top-ranked mutations characterize Omicron lineages BA.1, one of the first recognized recombinant lineages XB, Gamma P.1, and lineage JN.1.7, which had over 20% global prevalence in spring 2024. Half of these mutations have relatively high DMS reactivity for basepaired positions and half have very low DMS reactivity. By comparison, the synonymous substitution C29095U at an unpaired position has very high estimated mutational fitness and DMS reactivity. Despite having a higher median estimated mutational fitness (1.41 vs. 1.10), only 3 of the top 15 nonsynonymous C→U at unpaired positions define major lineages (BQ.1.1, JN.1.8.1, and BA.2.86.1).

Of particular note is C22227T at a basepaired position encoding the spike A222V mutation. This was one of the mutations that characterized the B.1.177 lineage. This lineage, also known as EU1, characterized a majority of sequences in Spain in summer 2020 and eventually in several other countries in Europe prior to the emergence of the Alpha variant. However, it was unclear whether this lineage had higher fitness than other lineages or if A222V specifically conferred a fitness advantage (Hodcroft et al., 2021). Further investigation as well as its recurrence in the major Delta sublineage AY.4.2 provided additional evidence of an increase in viral fitness and suggested molecular mechanisms (Ginex et al., 2022). Here, I focus on C→U substitutions for comparison to DMS reactivity data, but I also note that top-ranked G→U substitutions at basepaired positions are rich in mutations to ORF3a and also include mutations that characterize variants of concern, such as nucleocapsid D377Y in Delta. Lastly, note that, following the coarse adjustment for basepairing inferred from synonymous substitutions, non-synonymous C→U substitutions characterizing major variants now have some of the highest estimated mutational fitness for C→U substitutions (Figure 2, right).

Discussion

This analysis shows that it is informative to combine apparent viral fitness, inferred from millions of patient SARS-CoV-2 sequences published during the pandemic, with accurate secondary structure measurements. It is important to remember that apparent “mutational fitness” results from a combination of the rates at which diversity is generated and the subsequent selection processes. Importantly, genome secondary structure can impact both. However, even the unprecedented density of sampling SARS-CoV-2 genomes has been insufficient to reliably infer fitness impacts of single mutations more directly from dynamics subsequent to occurrences in the SARS-CoV-2 phylogenetic tree (Bloom and Neher, 2023; Richard et al., 2021).

Further investigation into phenomena reported here, such as the lack of apparent secondary structure dependence for A→G and G→A substitutions, could inform investigation of underlying mutation and selection mechanisms. An analysis focused on basepairing dependence of C→U substitution frequencies considered that this may be driven by conservation of secondary structure (Gout et al., 2024). However, the lack of basepairing dependence for A→G and G→A substitutions supports the hypothesis that this effect, on average, arises from differences in chemical reaction rates (Li et al., 2023). Nonetheless, future analyses could consider whether substitution rates have different basepairing dependencies in regions in which synonymous mutations exhibit fitness reductions (Bloom and Neher, 2023). The extents of enzymatic and non-enzymatic mutation processes driving differences in substitution rates remains unclear (Simmonds, 2024). Comparing the degree to which rates depend on secondary structure could provide insight to support different mutation mechanisms. A recent in vitro study suggested that oxidation of guanine can give rise to G→U substitutions (Akagawa et al., 2024), which like C→U substitutions, are both very common in human SARS-CoV-2 infections and exhibit strong basepairing dependence.

I suggest that secondary structure, along with other data correlating with substitution frequency, can be used to refine estimates of mutational fitness both for future SARS-CoV-2 variants and for other viruses. More sophisticated analysis can incorporate structural heterogeneity (Lan et al., 2022) as well as local sequence context (Lamb et al., 2024). Figure S3 shows how the coarse adjustment described in this paper does not capture all information available in DMS reactivity data, and DMS interrogates adenine and cytosine basepairing while SHAPE reagents can be used to investigate backbone flexibility for all nucleotides. Clearly, datasets from different experiments using different chemical probes can be combined to produce more accurate estimates of apparent mutational fitness. Furthermore, additional measurements of secondary structure for genomes of new variants or modeling may reveal significant changes to secondary structure since 2020. This could be evident in in vitro substitution rates for different SARS-CoV-2 variants (Gout et al., 2024). Lastly, secondary structures determined from in vitro chemical probing experiments may differ from secondary structures in vivo. For the spike protein, the correlation between estimated mutational fitness and pseudovirus entry quantified by deep mutational scanning serves as one metric that can be used to optimize models (Lamb et al., 2024). However, it is critical to evaluate uncertainty in any model estimating fitness of a new variant. Initial estimates can be refined by rapid in vitro characterization and continued genomic surveillance.